---
pagetitle: "Lesson 37"
title: "Hierarchical Clustering in R — Dendrograms, Linkage, Cutting at k, and Comparison with K-means"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Hierarchical Clustering

Hierarchical clustering produces a **nested sequence of partitions** represented as a dendrogram.  
In agglomerative clustering (bottom-up), each observation starts as its own cluster; clusters are iteratively merged according to a **linkage** rule and a distance measure.

Given a dissimilarity $d(\cdot,\cdot)$, common linkages include:

- **Complete**: $d(A,B) = \max_{i \in A, j \in B} d(i,j)$ (tends to compact clusters).
- **Average**: $d(A,B) = \frac{1}{|A||B|}\sum_{i \in A, j \in B} d(i,j)$ (balanced).
- **Ward**: merges that minimize the increase in within-cluster SSE (often strong baseline with Euclidean distance).

In this lesson:
- We build dendrograms with multiple linkages.
- We cut the tree at $k$.
- We validate clusters with silhouette.
- We compare against K-means using ARI and centroid profiles.

# Step 1: Import data (business dataset)

```{r}
library(readr)
library(dplyr)

local_path <- "raw_data/wholesale_customers.csv"

if (!file.exists(local_path)) {
  dir.create("raw_data", showWarnings = FALSE)
  download.file(url_path, local_path, mode = "wb")
}

df_raw <- read_csv(local_path, show_col_types = FALSE)

glimpse(df_raw)
```

# Step 2: Preprocessing (log-transform + scale)

Spending variables are typically right-skewed; a common preprocessing is $x \leftarrow \log(1+x)$ followed by standardization.

```{r}
spend_vars <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

df <- df_raw |>
  mutate(across(all_of(spend_vars), ~ log1p(.x)))

X <- df |>
  select(all_of(spend_vars)) |>
  mutate(across(everything(), scale)) |>
  as.data.frame()

summary(X)
```

# Step 3: Distances and dendrograms

```{r}
d <- dist(X, method = "euclidean")

hc_complete <- hclust(d, method = "complete")
hc_average  <- hclust(d, method = "average")
hc_ward     <- hclust(d, method = "ward.D2")

par(mfrow = c(1, 3))
plot(hc_complete, main = "Hierarchical (complete)", cex = 0.6)
plot(hc_average,  main = "Hierarchical (average)",  cex = 0.6)
plot(hc_ward,     main = "Hierarchical (ward.D2)",  cex = 0.6)
par(mfrow = c(1, 1))
```

Interpretation guideline:
- Very “chained” dendrograms can indicate poor separation or inappropriate linkage/distance.
- Ward often yields more balanced groups when Euclidean geometry is meaningful.

# Step 4: Cutting at k and silhouette validation

We choose a candidate range for $k$ and compute average silhouette.

```{r}
library(cluster)
library(ggplot2)

avg_sil <- function(labels, d) {
  s <- silhouette(labels, d)
  mean(s[, 3])
}

k_grid <- 2:12
sil_df <- data.frame(
  k = k_grid,
  sil_complete = NA_real_,
  sil_average  = NA_real_,
  sil_ward     = NA_real_
)

for (i in seq_along(k_grid)) {
  k <- k_grid[i]
  sil_df$sil_complete[i] <- avg_sil(cutree(hc_complete, k = k), d)
  sil_df$sil_average[i]  <- avg_sil(cutree(hc_average,  k = k), d)
  sil_df$sil_ward[i]     <- avg_sil(cutree(hc_ward,     k = k), d)
}

sil_long <- tidyr::pivot_longer(
  sil_df, cols = -k, names_to = "linkage", values_to = "avg_silhouette"
)

ggplot(sil_long, aes(x = k, y = avg_silhouette, color = linkage)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Average silhouette for hierarchical clustering",
    x = "Number of clusters (k)",
    y = "Average silhouette"
  )
```

Pick $k$ where silhouette is reasonably high and the solution is interpretable (avoid excessively tiny clusters unless expected).

# Step 5: Final hierarchical solution (example: Ward + chosen k)

```{r}
k_final <- sil_df$k[which.max(sil_df$sil_ward)]
cl_hc <- cutree(hc_ward, k = k_final)

table(cl_hc)
```

Silhouette plot for the final solution:

```{r}
sil_obj <- silhouette(cl_hc, d)
plot(sil_obj, border = NA, main = paste("Silhouette plot — Ward, k =", k_final))
summary(sil_obj)
```

# Step 6: Compare with K-means (same k)

K-means minimizes within-cluster SSE:
$ \text{WCSS}(k) = \sum_{r=1}^k \sum_{i \in C_r} \|x_i - \bar{x}_r\|^2 $.

```{r}
set.seed(123)
km <- kmeans(X, centers = k_final, nstart = 50)

table(km$cluster)
```

Quantitative comparison (Adjusted Rand Index):

```{r}
library(mclust)
ari_hc_km <- adjustedRandIndex(cl_hc, km$cluster)
ari_hc_km
```

Interpretation:
- ARI close to 0.65: good agreement between partitions.
- Lower ARI: methods capture different structure; investigate cluster shapes, outliers, and linkage choice.

# Step 7: Profile clusters (centroid heatmap)

```{r}
library(tidyr)

centers_hc <- aggregate(X, by = list(cluster = cl_hc), FUN = mean)
centers_long <- centers_hc |>
  pivot_longer(-cluster, names_to = "feature", values_to = "z")

ggplot(centers_long, aes(x = feature, y = factor(cluster), fill = z)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Hierarchical cluster profiles (z-scores)", x = NULL, y = "Cluster")
```

# Summary

- Hierarchical clustering provides a dendrogram and supports **post-hoc** selection of $k$ by cutting the tree.
- Linkage choice strongly affects results; Ward is often a strong baseline with Euclidean distances.
- Silhouette helps select $k$ and assess separation.
- Comparing with K-means via ARI and centroid profiles clarifies when the two approaches agree or diverge.
