---
pagetitle: "Lesson 36"
title: "Ridge & Lasso Regression in R — Penalized Linear Models with glmnet"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Ridge & Lasso Regression (Penalized OLS)

## Why penalization?

OLS estimates can become unstable when:

- predictors are highly correlated (multicollinearity),
- the number of predictors is large relative to the sample size,
- you want a model that generalizes better (lower variance) and/or performs variable selection.

Penalized regression adds a constraint/penalty to shrink coefficients toward zero, trading a bit of bias for lower variance.

***

## Theory: Ridge, Lasso, Elastic Net

Assume a linear model $$y = X\beta + \epsilon$$ with $$X \in \mathbb{R}^{n \times p}$$.
The OLS objective minimizes RSS:
$$
\min_{\beta_0,\beta}\ \frac{1}{2n}\sum_{i=1}^n (y_i - \beta_0 - x_i^T\beta)^2
$$

### Ridge (L2 penalty)

$$
\min_{\beta_0,\beta}\ \frac{1}{2n}\sum_{i=1}^n (y_i - \beta_0 - x_i^T\beta)^2\ +\ \lambda \sum_{j=1}^p \beta_j^2
$$

- Shrinks coefficients smoothly.
- Typically **does not** set coefficients exactly to zero (no hard feature selection).
- Very useful under multicollinearity.

### Lasso (L1 penalty)

$$
\min_{\beta_0,\beta}\ \frac{1}{2n}\sum_{i=1}^n (y_i - \beta_0 - x_i^T\beta)^2\ +\ \lambda \sum_{j=1}^p |\beta_j|
$$

- Encourages **sparsity**: some $\hat\beta_j = 0$.
- Performs variable selection (often easier interpretation).

### Elastic Net (mix L1 + L2)

Elastic Net combines both:
$$
\min_{\beta_0,\beta}\ \frac{1}{2n}\|y-\beta_0\mathbf{1}-X\beta\|_2^2
\ +\ \lambda\left[(1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha\|\beta\|_1\right]
$$
with $0 \le \alpha \le 1$:

- $\alpha = 0$ → ridge
- $\alpha = 1$ → lasso
- $0<\alpha<1$ → elastic net

### Hyperparameter intuition

- $\lambda = 0$: no penalization (≈ OLS).
- $\lambda \uparrow$: stronger shrinkage; lasso sets more coefficients to zero.
- $\alpha$: controls **sparsity** vs **stability**.

***

## Practice in R with glmnet

Key points:

- `glmnet` expects a numeric matrix `x` (use `model.matrix()` for factors).
- Use cross-validation to select $\lambda$ via `cv.glmnet()`.
- Common choices:
  - `lambda.min`: minimizes CV error.
  - `lambda.1se`: simplest model within 1 standard error (more regularization; often more robust).

***

# Example: Ridge vs Lasso on mtcars (Gaussian regression)

We predict `mpg` using a mix of numeric and categorical features.

## Step 1: Data preparation (design matrix)

```{r}
library(dplyr)

df <- mtcars |>
  mutate(
    cyl = factor(cyl),
    am  = factor(am, labels = c("auto", "manual")),
    vs  = factor(vs)
  )

# Model matrix: adds intercept column by default; glmnet handles intercept separately,
# so we remove it with "-1" here and let glmnet fit intercept = TRUE.
X <- model.matrix(mpg ~ . - 1, data = df)
y <- df$mpg

dim(X)
head(X[, 1:6])
```

## Step 2: Train/test split

```{r}
set.seed(123)
n <- nrow(X)
idx_train <- sample.int(n, size = floor(0.75 * n))

X_train <- X[idx_train, , drop = FALSE]
y_train <- y[idx_train]

X_test <- X[-idx_train, , drop = FALSE]
y_test <- y[-idx_train]
```

## Step 3: Baseline OLS (for comparison)

```{r}
train_ols <- data.frame(mpg = y_train, X_train)
test_ols  <- data.frame(mpg = y_test,  X_test)

fit_ols <- lm(mpg ~ ., data = train_ols)
summary(fit_ols)

pred_ols <- predict(fit_ols, newdata = test_ols)
rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

rmse_ols <- rmse(y_test, pred_ols)
rmse_ols
```

***

## Step 4: Ridge regression (alpha = 0) with CV

```{r}
# install.packages("glmnet")  # if needed
library(glmnet)

set.seed(123)
cv_ridge <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 0,            # ridge
  nfolds = 10,          # default is 10
  standardize = TRUE
)

plot(cv_ridge)
cv_ridge$lambda.min
cv_ridge$lambda.1se
```

Fit ridge at selected lambda and evaluate:

```{r}
pred_ridge_min <- predict(cv_ridge, newx = X_test, s = "lambda.min")
pred_ridge_1se <- predict(cv_ridge, newx = X_test, s = "lambda.1se")

rmse_ridge_min <- rmse(y_test, pred_ridge_min)
rmse_ridge_1se <- rmse(y_test, pred_ridge_1se)

c(rmse_ols = rmse_ols, rmse_ridge_min = rmse_ridge_min, rmse_ridge_1se = rmse_ridge_1se)
```

Inspect coefficients:

```{r}
coef(cv_ridge, s = "lambda.min")[1:10, , drop = FALSE]
```

***

## Step 5: Lasso regression (alpha = 1) with CV

```{r}
set.seed(123)
cv_lasso <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 1,            # lasso
  nfolds = 10,
  standardize = TRUE
)

plot(cv_lasso)
cv_lasso$lambda.min
cv_lasso$lambda.1se
```

Evaluate and compare:

```{r}
pred_lasso_min <- predict(cv_lasso, newx = X_test, s = "lambda.min")
pred_lasso_1se <- predict(cv_lasso, newx = X_test, s = "lambda.1se")

rmse_lasso_min <- rmse(y_test, pred_lasso_min)
rmse_lasso_1se <- rmse(y_test, pred_lasso_1se)

c(
  rmse_ols = rmse_ols,
  rmse_lasso_min = rmse_lasso_min,
  rmse_lasso_1se = rmse_lasso_1se
)
```

Check sparsity (number of non-zero coefficients):

```{r}
b_lasso_min <- coef(cv_lasso, s = "lambda.min")
nnz <- sum(b_lasso_min != 0)
nnz
```

List selected predictors:

```{r}
selected <- rownames(b_lasso_min)[as.vector(b_lasso_min != 0)]
selected
```

***

## Step 6: Coefficient paths (optional)

```{r}
fit_lasso_path <- glmnet(X_train, y_train, alpha = 1, standardize = TRUE)
plot(fit_lasso_path, xvar = "lambda", label = TRUE, main = "Lasso coefficient paths")
abline(v = -log(cv_lasso$lambda.min), lty = 2, col = "red")

fit_ridge_path <- glmnet(X_train, y_train, alpha = 0, standardize = TRUE)
plot(fit_ridge_path, xvar = "lambda", label = FALSE, main = "Ridge coefficient paths")
abline(v = -log(cv_ridge$lambda.min), lty = 2, col = "red")
```

***

## Optional: Elastic Net (alpha between 0 and 1)

If predictors are highly correlated, elastic net can outperform pure lasso.

```{r}
set.seed(123)
cv_enet <- cv.glmnet(
  x = X_train,
  y = y_train,
  alpha = 0.5,          # elastic net
  nfolds = 10,
  standardize = TRUE
)

plot(cv_enet)
pred_enet <- predict(cv_enet, newx = X_test, s = "lambda.min")
rmse_enet <- rmse(y_test, pred_enet)

c(rmse_lasso_min = rmse_lasso_min, rmse_ridge_min = rmse_ridge_min, rmse_enet_min = rmse_enet)
```

***

# Best practices & warnings

- Always use **cross-validation** for $\lambda$; prefer `lambda.min` when you want stability and parsimony.
- Standardization is important so that the penalty treats predictors comparably.
- Lasso can be unstable when predictors are strongly correlated (it may pick one arbitrarily); elastic net is often safer.
- If the goal is inference (p-values, CI), penalized regression needs extra care (post-selection inference); treat as predictive by default.

***

# Summary

You learned how to:
- Define ridge/lasso/elastic net with penalized objectives.
- Fit ridge and lasso in R using `glmnet` + `cv.glmnet()`.
- Select $\lambda$ via CV (`lambda.min` vs `lambda.1se`).
- Compare test RMSE and interpret shrinkage vs sparsity.

Penalized regression is a modern alternative to stepwise selection for regression problems with many correlated predictors.
