---
pagetitle: "Lesson 39"
title: "Model-based Clustering in R — Gaussian Mixtures, Soft Assignments, and Component Selection"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Model-based clustering (Gaussian Mixture Models)

Model-based clustering assumes data arise from a mixture distribution:
$ p(x) = \sum_{g=1}^G \pi_g \, \phi(x \mid \mu_g, \Sigma_g) $,
where $\pi_g \ge 0$ and $\sum_g \pi_g = 1$.

Key properties:

- **Soft assignments** via posterior probabilities:
$ \tau_{ig} = P(z_i=g \mid x_i) $.
- Clusters can be **elliptical** (through $\Sigma_g$), unlike K-means which is inherently spherical in Euclidean space.
- The number of components $G$ can be selected using information criteria (commonly BIC).

In this lesson:

- Fit a GMM with automatic model selection.
- Inspect soft assignments (uncertainty).
- Compare with K-means.

# Step 1: Import + preprocessing

```{r}
library(readr)
library(dplyr)

local_path <- "raw_data/wholesale_customers.csv"
df_raw <- read_csv(local_path, show_col_types = FALSE)

spend_vars <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

X <- df_raw |>
  mutate(across(all_of(spend_vars), ~ log1p(.x))) |>
  select(all_of(spend_vars)) |>
  mutate(across(everything(), scale)) |>
  as.matrix()
```

# Step 2: Fit model-based clustering and select G by BIC

```{r}
# install.packages("mclust")  # if needed
library(mclust)

set.seed(123)
fit <- Mclust(X, G = 1:12)

fit
plot(fit, what = "BIC")
```

Commentary:

- If the BIC curve has a clear maximum, the selected $G$ is usually stable.
- If BIC is flat across many $G$, clustering structure may be weak or sensitive to preprocessing.

# Step 3: Hard vs soft clustering outputs

```{r}
cl <- fit$classification          # hard labels
prob <- fit$z                     # posterior probabilities (n x G)
uncert <- 1 - apply(prob, 1, max) # simple uncertainty measure

table(cl)
summary(uncert)
```

High uncertainty points indicate observations not clearly explained by any single component.

# Step 4: Visualization (PCA + uncertainty)

```{r}
library(ggplot2)

pc <- prcomp(X)
plot_df <- data.frame(
  PC1 = pc$x[, 1],
  PC2 = pc$x[, 2],
  cluster = factor(cl),
  uncertainty = uncert
)

ggplot(plot_df, aes(PC1, PC2, color = cluster)) +
  geom_point(aes(alpha = 1 - uncertainty), size = 2) +
  theme_minimal() +
  labs(
    title = "Model-based clustering (GMM) — PCA projection",
    subtitle = "Point transparency reflects assignment certainty",
    x = "PC1", y = "PC2"
  )
```

# Step 5: Compare with K-means (same number of clusters)

```{r}
set.seed(123)
km <- kmeans(X, centers = length(unique(cl)), nstart = 50)

ari <- adjustedRandIndex(cl, km$cluster)
ari
```

Interpretation:

- If ARI is high, both methods recover similar structure.
- If ARI is lower, investigate covariance shapes: GMM can fit ellipses; K-means cannot.

# Step 6: Cluster profiling

```{r}
df_prof <- as.data.frame(X)
df_prof$cluster <- factor(cl)

profiles <- df_prof |>
  group_by(cluster) |>
  summarize(across(where(is.numeric), mean), .groups = "drop")

profiles
```

For business communication, it is often preferable to profile clusters on **original-scale** variables (before scaling/log), and report relative differences.

# Summary

- Model-based clustering provides a probabilistic framework with soft assignments.
- It naturally supports elliptical clusters and explicit uncertainty quantification.
- BIC-based selection helps choose the number of components.
- Comparing GMM vs K-means clarifies whether “spherical” clustering is too restrictive.
