---
pagetitle: "Lesson 2"
title: "R Data Types, Arithmetic, and Logical Operators (with Examples)"
# author: "Dott. Gianluca Sottile"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

<br>

# Basic data types {#buttons .tabset .tabset-fade .tabset-pills}
***

R works with several data types and data structures, including:

+ Scalars
+ Vectors (numeric, character, logical)
+ Matrices
+ Data frames
+ Lists

__Basic types__

+ `4.5` is a decimal number (type: numeric).
+ `4L` is an integer (type: integer). Integers are a subset of numeric values.
+ `TRUE` or `FALSE` are Boolean values (type: logical).
+ Values inside `" "` or `' '` are text (strings). They are stored as type: character.

You can check the type of an object with `class()`.

## Example 1:
```{r}
# Numeric
x <- 28
class(x)
```

## Example 2:
```{r}
# Character
y <- "R is fantastic"
class(y)
```

## Example 3:
```{r}
# Logical
z <- TRUE
class(z)
```

## Example 4 (optional):
```{r}
# Integer (note the L suffix)
w <- 4L
class(w)
```

# Variables {#buttons1 .tabset .tabset-fade .tabset-pills}
***

Variables store values and are a key component in programming. In R, a variable can store almost anything: a number, an object, a statistical result, a vector, a dataset, a model, or predictions. You can reuse a variable later by typing its name.

To create a variable, you assign a value to a name. Variable names should not contain spaces; you can use `_` to separate words.

To assign a value, use `<-` (recommended) or `=`.

Here is the syntax:

```{r, eval=FALSE}
# Recommended assignment operator
name_of_variable <- value

# Also used sometimes (e.g., inside function calls)
name_of_variable = value
```

In the console, you can run the following examples:

## Example 1:
```{r}
# Create a variable
x <- 42
x
```

## Example 2:
```{r}
y <- 10
y
```

## Example 3:
```{r}
# Use variables in an expression
x - y
```

# Vectors {#buttons2 .tabset .tabset-fade .tabset-pills}
***

A vector is a one-dimensional sequence of values. You can create a vector from any basic type. The simplest way to build a vector is using `c()`.

## Example 1:
```{r}
# Numeric
vec_num <- c(1, 10, 49)
vec_num
```

## Example 2:
```{r}
# Character
vec_chr <- c("a", "b", "c")
vec_chr
```

## Example 3:
```{r}
# Logical
vec_lgl <- c(TRUE, FALSE, TRUE)
vec_lgl
```

## Example 4:
You can perform arithmetic operations on vectors element-by-element.

```{r}
vect_1 <- c(1, 3, 5)
vect_2 <- c(2, 4, 6)

sum_vect <- vect_1 + vect_2
sum_vect
```

## Example 5:
In R, you can slice a vector using square brackets `[]`. For example, `x[1:5]` selects elements 1 to 5.

```{r}
slice_vector <- 1:10
slice_vector[1:5]
```

## Example 6:
A quick way to create a sequence is `:`.

```{r}
1:10
```

# Arithmetic Operators {#buttons3 .tabset .tabset-fade .tabset-pills}
***

Let’s look at basic arithmetic operators in R:

```{r, echo=FALSE}
ops_arith <- data.frame(
  Operator = c("+", "-", "*", "/", "^", "%%"),
  Description = c(
    "Addition",
    "Subtraction",
    "Multiplication",
    "Division",
    "Exponentiation",
    "Modulo (remainder)"
  )
)

knitr::kable(ops_arith) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

Notes:

- Lines starting with `#` are comments; R ignores them.
- In rendered R Markdown, printed output is often shown with a `##` prefix.

## Example 1:
```{r}
# Addition
3 + 4
```

## Example 2:
```{r}
# Multiplication
3 * 5
```

## Example 3:
```{r}
# Division
(5 + 5) / 2
```

## Example 4:
```{r}
# Exponentiation
2^5
```

## Example 5:
```{r}
# Modulo
28%% 6
```

# Logical Operators {#buttons4 .tabset .tabset-fade .tabset-pills}
***

Logical operators return `TRUE` or `FALSE`. They are used to test conditions and filter values.

```{r, echo=FALSE}
ops_logical <- data.frame(
  Operator = c("<", "<=", ">", ">=", "==", "!=", "!", "x & y", "x | y", "isTRUE(x)"),
  Description = c(
    "Less than",
    "Less than or equal to",
    "Greater than",
    "Greater than or equal to",
    "Exactly equal to",
    "Not equal to",
    "Not x",
    "x AND y (vectorised)",
    "x OR y (vectorised)",
    "Test whether x is TRUE"
  )
)

knitr::kable(ops_logical) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

To select values that satisfy a condition, use:

```{r, eval=FALSE}
variable_name[condition]
```

## Example 1:
```{r}
# Create a vector from 1 to 10
logical_vector <- 1:10

# Test a condition (returns TRUE/FALSE for each element)
logical_vector > 5
```

In the output above, R compares each element to the condition `logical_vector > 5`. If the element is strictly greater than 5, the result is `TRUE`; otherwise it is `FALSE`.

## Example 2:
To extract only the values that meet the condition, put the condition inside `[]` after the vector.

```{r}
logical_vector[logical_vector > 5]
```

## Example 3:
You can combine conditions with `&` (AND) and use parentheses for clarity.

```{r}
logical_vector[(logical_vector > 4) & (logical_vector < 7)]
```
