---
pagetitle: "Lesson 4"
title: "Factors in R: Categorical vs Continuous Variables"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# What is a factor in R?

A **factor** is R’s data type for **categorical variables** (variables that take a limited set of values, called *levels*). [web:833]  
In most datasets you will see two broad types of variables:

- **Categorical**: a finite set of categories (e.g., country, gender, job title).
- **Continuous**: numeric measurements on a scale (e.g., income, temperature, price).

R models and plotting functions often treat factors differently from numeric variables, which is why it matters to store categorical data as factors. [web:837]

# Categorical variables

You create factors with `factor()`:

```{r, eval=FALSE}
factor(x, levels = NULL, labels = levels, ordered = FALSE)
```

Key arguments: [web:833]
- `x`: a vector (typically character, integer, or numeric codes).
- `levels`: the set of possible values (and their order, if provided).
- `labels`: the human-readable labels for the levels.
- `ordered`: whether levels should be treated as ordered.

## Example: character to factor

```{r}
gender <- c("Male", "Female", "Female", "Male", "Male")

class(gender)
gender_f <- factor(gender)
class(gender_f)

gender_f
levels(gender_f)
```

## Example: numeric codes with labels

```{r}
gender_code <- c(1, 0, 0, 1, 1)

gender_labeled <- factor(
  gender_code,
  levels = c(0, 1),
  labels = c("Female", "Male")
)

gender_labeled
```

# Nominal vs ordinal categorical variables

## Nominal factors

Nominal categories have **no natural order** (e.g., colors).

```{r}
color <- c("blue", "red", "green", "white", "black", "yellow")
color_f <- factor(color)
color_f
levels(color_f)
```

The level order is just how R stores it (often alphabetical unless you specify levels explicitly). [web:837]

## Ordinal factors {#buttons .tabset .tabset-fade .tabset-pills}

Ordinal categories have a **natural order** (e.g., low < medium < high).  
In R, you can encode that by setting `ordered = TRUE` and supplying levels in the correct order. [web:833][web:844]

### Example 1: create an ordered factor

```{r}
day <- c("evening", "morning", "afternoon", "midday", "midnight", "evening")

day_ord <- factor(
  day,
  ordered = TRUE,
  levels = c("morning", "midday", "afternoon", "evening", "midnight")
)

day_ord
```

### Example 2: summarize levels

```{r}
summary(day_ord)
```

Because the factor is ordered, printing shows the order of levels using `<`. [web:837]

# Continuous variables

Continuous variables are usually stored as `numeric` (or sometimes `integer`). For example, `mtcars$mpg` is numeric, so it is a continuous variable.

```{r}
dataset <- mtcars
class(dataset$mpg)
```

# Practical tip (modeling)

In many modeling functions, factors are automatically expanded into indicator (dummy) variables in the design matrix. This is one reason factors are important for regression and other ML methods. [web:839]
