---
pagetitle: "Lesson 38"
title: "DBSCAN in R — eps, minPts, Noise Handling, and Non-spherical Clusters"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# DBSCAN (Density-Based Clustering)

DBSCAN groups points based on density rather than distance to centroids. Key concepts:

- $\varepsilon$-neighborhood: $N_\varepsilon(x) = \{y: \|x-y\| \le \varepsilon\}$.
- **Core point**: $|N_\varepsilon(x)| \ge \text{minPts}$.
- **Border point**: not core but density-reachable from a core point.
- **Noise**: points not density-reachable from any core point.

DBSCAN can detect **non-spherical** clusters and identifies **outliers/noise** explicitly.

In this lesson

- We preprocess business data as in previous lesson.
- We choose `eps` using kNN distance plot.
- We run DBSCAN and interpret clusters vs noise.
- We compare DBSCAN with K-means on the non-noise subset.

# Step 1: Import + preprocessing

```{r}
library(readr)
library(dplyr)

local_path <- "raw_data/wholesale_customers.csv"

if (!file.exists(local_path)) {
  dir.create("raw_data", showWarnings = FALSE)
  download.file(url_path, local_path, mode = "wb")
}

df_raw <- read_csv(local_path, show_col_types = FALSE)

spend_vars <- c("Fresh", "Milk", "Grocery", "Frozen", "Detergents_Paper", "Delicassen")

X <- df_raw |>
  mutate(across(all_of(spend_vars), ~ log1p(.x))) |>
  select(all_of(spend_vars)) |>
  mutate(across(everything(), scale)) |>
  as.data.frame()
```

# Step 2: Parameter selection (minPts and eps)

A common heuristic sets $\text{minPts} \approx 2p$ where $p$ is the number of features.

```{r}
p <- ncol(X)
minPts <- max(5, 2 * p)
minPts
```

To choose `eps`, inspect the kNN distance plot: pick `eps` near the “knee”.

```{r}
library(dbscan)

kNNdistplot(X, k = minPts)
abline(h = 1, col = "red", lty = 2)
```

Replace the horizontal line with your candidate value for `eps` after inspecting the knee.

# Step 3: Fit DBSCAN

```{r}
eps <- 1.0  # adjust after inspecting the kNN plot

db <- dbscan(X, eps = eps, minPts = minPts)
db

table(db$cluster)  # 0 = noise
```

Interpretation:
- If most points are noise, `eps` may be too small or minPts too large.
- If almost no noise and very few clusters, `eps` may be too large (merging structures).

# Step 4: Visualization (PCA projection)

```{r}
library(ggplot2)

pc <- prcomp(X)
plot_df <- data.frame(
  PC1 = pc$x[, 1],
  PC2 = pc$x[, 2],
  cluster = factor(db$cluster)
)

ggplot(plot_df, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.8, size = 2) +
  theme_minimal() +
  labs(
    title = "DBSCAN clusters (PCA projection)",
    subtitle = "Cluster 0 denotes noise",
    x = "PC1", y = "PC2"
  )
```

# Step 5: Internal validation (silhouette on non-noise points)

Silhouette is not directly defined for noise. A practical approach is to compute silhouette on the subset with cluster label $\ge 1$.

```{r}
library(cluster)

idx <- which(db$cluster > 0)
X_nonoise <- X[idx, , drop = FALSE]
cl_nonoise <- db$cluster[idx]

if (length(unique(cl_nonoise)) >= 2) {
  d <- dist(X_nonoise)
  sil <- silhouette(cl_nonoise, d)
  summary(sil)
  plot(sil, border = NA, main = "Silhouette (non-noise subset)")
}
```

# Step 6: Comparison with K-means (on non-noise subset)

```{r}
library(mclust)

k_db <- length(unique(cl_nonoise))
set.seed(123)
km <- kmeans(X_nonoise, centers = k_db, nstart = 50)

ari <- adjustedRandIndex(cl_nonoise, km$cluster)
ari
```

Interpretation:

- Agreement is not expected to be perfect, because DBSCAN optimizes density connectivity, not SSE.
- Differences often highlight **non-spherical** structure or outliers.

# Step 7: Practical reporting

For a business segmentation report:

- Report number of clusters found (excluding noise) and proportion of noise.
- Provide cluster profiles (feature means in original units, not only z-scores).
- Explain that DBSCAN can isolate atypical customers as noise/outliers, which may be a feature rather than a drawback.

# Summary

- DBSCAN is density-based and can recover non-spherical clusters.
- `minPts` and `eps` are critical; the kNN distance plot is a practical tool for selecting `eps`.
- Noise labeling provides a principled way to treat outliers.
- Compare with K-means on the non-noise subset to understand structural differences.
