---
pagetitle: "Lesson 7"
title: "R Sort a Data Frame using Order()"
# author: "Dott. Gianluca Sottile"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

<br>

In data analysis you can sort your data according to a certain variable in the dataset. In R, we can use the help of the function order(). In R, we can easily sort a vector of continuous variable or factor variable. Arranging the data can be of ascending or descending order.

__Syntax:__

```{r, dont-eval, eval=FALSE}
sort(x, decreasing = FALSE, na.last = TRUE)
```

__Argument:__

+ x: A vector containing continuous or factor variable
+ decreasing: Control for the order of the sort method. By default, decreasing is set to `FALSE`.
+ last: Indicates whether the `NA` 's value should be put last or not

__Example 1__
For instance, we can create a tibble data frame and sort one or multiple variables. A tibble data frame is a new approach to data frame. It improves the syntax of data frame and avoid frustrating data type formatting, especially for character to factor. It is also a convenient way to create a data frame by hand, which is our purpose here. To learn more about tibble, please refer to the vignette: <https://cran.r-project.org/web/packages/tibble/vignettes/tibble.html>

```{r}
library(dplyr)
set.seed(1234)
data_frame <- tibble(  
	c1 = rnorm(50, 5, 1.5),   
	c2 = rnorm(50, 5, 1.5),  
	c3 = rnorm(50, 5, 1.5),
	c4 = rnorm(50, 5, 1.5), 	
	c5 = rnorm(50, 5, 1.5)
)
# Sort by c1
df <- data_frame[order(data_frame$c1), ]
head(df)
```

__Example2__

```{r}
# Sort by c3 and c4
df <- data_frame[order(data_frame$c3, data_frame$c4), ]
head(df)
```

__Example 3__

```{r}
# Sort by c3(descending) and c4(acending)
df <- data_frame[order(-data_frame$c3, data_frame$c4), ]
head(df)
```

