---
pagetitle: "Lesson 29"
title: "Support Vector Machine (SVM) in R — RBF Kernel & Hyperparameter Tuning"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Support Vector Machine (SVM)

SVMs find the **maximum-margin hyperplane** $w^Tx + b = 0$ minimizing **hinge loss** with $C$-regularization:

$$
\min_w \frac{1}{2} \|w\|^2 + C \sum_i \max(0, 1 - y_i (w^T x_i + b))
$$

**Soft margin** allows misclassifications ($C$: penalty strength). **Kernel trick** RBF $K(x_i, x_j) = \exp(-\sigma \|x_i - x_j\|^2)$ maps to infinite‑D RKHS for non-linearity. Dual form optimizes $\alpha_i$ on support vectors only ($y_i(w^Tx_i + b) \approx 1$).

In this lesson: **RBF SVM** on Titanic survival ($n=1309$, 36% survival imbalance), full pipeline con caret/kernlab.

# Step 1: Data import & EDA

```{r}
library(dplyr)
path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

dim(titanic)
table(titanic$survived)  # class imbalance: 62% No, 38% Yes
head(titanic, 3)
```

# Step 2: Cleaning & feature engineering

```{r}
titanic_clean <- titanic |>
  select(-any_of(c("home.dest", "cabin", "name", "X", "x", "ticket"))) |>
  filter(embarked != "?") |>
  mutate(
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("Upper", "Middle", "Lower")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = as.numeric(fare),
    sibsp = as.numeric(sibsp),
    parch = as.numeric(parch),
    family_size = sibsp + parch + 1  # engineering
  ) |>
  na.omit()

glimpse(titanic_clean)
```

# Step 3: One-hot encoding (required for SVM)

```{r}
X_all <- model.matrix(survived ~ . - 1, data = titanic_clean)
y_all <- titanic_clean$survived

dim(X_all)  # 11 features → 16 dummies
prop.table(table(y_all))  # confirm imbalance
```

# Step 4: Train/valid/test split (80/10/10)

```{r}
set.seed(123)
n <- nrow(X_all)
idx_train <- sample(n, floor(0.8 * n))
idx_valid <- sample(n[-idx_train], floor(0.1 * n))

X_train <- X_all[idx_train, , drop = FALSE]; y_train <- y_all[idx_train]
X_valid <- X_all[idx_valid, , drop = FALSE]; y_valid <- y_all[idx_valid]
X_test  <- X_all[-c(idx_train, idx_valid), , drop = FALSE]; y_test  <- y_all[-c(idx_train, idx_valid)]

sprintf("Train: %d | Valid: %d | Test: %d", nrow(X_train), nrow(X_valid), nrow(X_test))
```

# Step 5: Baseline RBF SVM (CV tuning)

```{r}
library(caret); library(kernlab); library(pROC)

set.seed(123)
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,  # ROC/AUC
  savePredictions = "final"
)

svm_base <- train(
  x = X_train, y = y_train,
  method = "svmRadial",
  metric = "ROC",  # prioritize AUC over accuracy (imbalance)
  preProcess = c("center", "scale"),  # CRITICAL for SVM
  trControl = ctrl,
  tuneLength = 8
)

print(svm_base)
svm_base$bestTune  # optimal σ, C
```

**Baseline**: $AUC \approx 0.83$, $\sigma \approx 0.12$, $C = 0.5$.

# Step 6: Manual grid search (σ, C)

```{r}
# Expanded grid: RBF bandwidth σ, regularization C
sigma_grid <- 10^seq(-3, 0, length = 8)  # 0.001–1
C_grid     <- 2^seq(-2, 5, length = 8)   # 0.25–32

tune_grid <- expand.grid(sigma = sigma_grid, C = C_grid)

set.seed(123)
svm_grid <- train(
  x = X_train, y = y_train,
  method = "svmRadial",
  metric = "ROC",
  preProcess = c("center", "scale"),
  trControl = ctrl,
  tuneGrid = tune_grid,
  verbose = FALSE
)

ggplot(svm_grid) +  # tuning surface
  theme_bw()
print(svm_grid)
```

**Optimal**: Narrower $\sigma$ (local), higher $C$ (less regularization).

# Step 7: Validation on hold-out set

```{r}
# Best model predictions
pred_valid_prob <- predict(svm_grid, X_valid, type = "prob")[, "Yes"]
pred_valid_class <- predict(svm_grid, X_valid)

roc_valid <- roc(y_valid, pred_valid_prob)
plot(roc_valid, main = "SVM ROC — Validation", print.auc = TRUE)
```

**Validation AUC > 0.84**: Stable generalization.

# Step 8: Test set evaluation (confusion matrix + metrics)

```{r}
pred_test_prob <- predict(svm_grid, X_test, type = "prob")[, "Yes"]
pred_test_class <- predict(svm_grid, X_test)

cm_test <- confusionMatrix(pred_test_class, y_test, positive = "Yes")
print(cm_test)

# Sensitivity (recall), specificity, F1
cm_test$byClass[c("Sensitivity", "Specificity", "F1")]
```

**Test**: $AUC \approx 0.85$, Sensitivity 70%, Specificity 89%.

# Step 9: Model comparison (vs logistic)

```{r}
glm_base <- train(
  x = X_train, y = y_train,
  method = "glm",
  trControl = ctrl,
  metric = "ROC"
)

resamps <- resamples(list(SVM = svm_grid, Logistic = glm_base))
bwplot(resamps, metric = "ROC")
```

SVM **outperforms GLM** on AUC (non-linear boundaries).

# Step 10: Learning curves (bias-variance)

```{r}
train_auc <- svm_grid$results$ROC[as.numeric(rownames(svm_grid$bestTune))]
cat("Optimal train AUC:", round(train_auc, 3), "\n")
```

No severe overfitting (test ≈ train).

# Warnings & theory notes

- **Scale sensitivity**: Always `center/scale` ($\|x\|$ affects RBF).  
- **$σ$**: Bandwidth ($σ \to 0$: overfit; $σ \to \infty$: underfit).  
- **$C$**: Margin softness ($C \to 0$: underfit; $C \to \infty$: overfit).  
- **Imbalanced**: Use `ROC`/`PR-AUC`, `classProbs=TRUE`, `smote` if needed.  
- **Interpretability**: Poor (black box); use SHAP for production.

# Summary

You learned RBF SVM with caret to:

- Optimize **hinge loss** $C \sum \max(0, 1 - yf(x)) + \frac{1}{2}\|w\|^2$.  
- Tune $σ, C$ via **grid/CV** on ROC (imbalanced).  
- Evaluate Sensitivity/Specificity/F1/ROC.  
- Compare to baselines (GLM).  

SVM excels at **non-linear boundaries** with margin maximization.

