---
title: "Simple, Multiple Linear and Stepwise Regression (with Example)"
pagetitle: "Lesson 23"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Linear regression (OLS): theory + practice

Linear regression models the conditional mean of a response as a linear function of predictors: $$ \mathbb{E}[Y \mid X] = X\beta $$.  
Under the classical assumptions (linearity in parameters, zero-mean errors conditional on $$X$$, homoscedasticity, no perfect multicollinearity), OLS is BLUE: **best linear unbiased estimator**.

We cover:
- Simple regression (by hand + `lm()`).
- Multiple regression (diagnostics, collinearity, inference).
- Regression with factors (dummy coding, reference levels).
- Stepwise regression (AIC) + **out-of-sample validation**.

***

## 1) Simple linear regression

Model:
$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i, \quad \mathbb{E}[\epsilon_i \mid x_i] = 0
$$

### Data + scatterplot

```{r}
library(ggplot2)
library(readr)

women_df <- read_csv("raw_data/women.csv", show_col_types = FALSE)

ggplot(women_df, aes(x = height, y = weight)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  theme_classic() +
  labs(title = "Women dataset: height vs weight")
```

### OLS estimates (by hand)

In simple regression:
$$
\hat{\beta}_1 = \frac{\text{Cov}(x,y)}{\text{Var}(x)}, \quad
\hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x}
$$

```{r}
beta_1 <- cov(women_df$height, women_df$weight) / var(women_df$height)
beta_0 <- mean(women_df$weight) - beta_1 * mean(women_df$height)
c(beta_0 = beta_0, beta_1 = beta_1)
```

### Fit with `lm()` + interpretation

```{r}
fit_simple <- lm(weight ~ height, data = women_df)
summary(fit_simple)
coef(fit_simple)
```

### Confidence interval vs prediction interval

A confidence interval targets the **mean response** $$ \mathbb{E}[Y \mid X=x] $$, while a prediction interval targets a **new observation** $$Y_{\text{new}}$$, and is therefore typically wider.

```{r}
new_h <- data.frame(height = c(60, 65, 70))

pred_ci <- predict(fit_simple, newdata = new_h, interval = "confidence", level = 0.95)
pred_pi <- predict(fit_simple, newdata = new_h, interval = "prediction", level = 0.95)

cbind(new_h, round(pred_ci, 2), round(pred_pi, 2))
```

***

## 2) Multiple linear regression

Model:
$$
y = \beta_0 + \beta_1 x_1 + \cdots + \beta_k x_k + \epsilon,
\quad \hat{\beta} = (X^T X)^{-1} X^T y
$$

We use `mtcars` to predict `mpg`.

### Fit a baseline model

```{r}
library(dplyr)

cars_cont <- mtcars |>
  select(mpg, disp, hp, drat, wt)

fit_multi <- lm(mpg ~ disp + hp + drat + wt, data = cars_cont)
summary(fit_multi)
anova(fit_multi)
```

### Diagnostics (linearity, normality, leverage)

```{r}
par(mfrow = c(2, 2))
plot(fit_multi)
par(mfrow = c(1, 1))
```

### Influential points (Cook’s distance)

Cook’s distance measures how much the fitted model would change if an observation were removed, combining residual size and leverage.

```{r}
cd <- cooks.distance(fit_multi)

plot(cd, type = "h", main = "Cook's distance (mtcars)", ylab = "Cook's D")
abline(h = 0, col = "grey60")

# Show the most influential points
head(sort(cd, decreasing = TRUE), 5)
```

### Multicollinearity (VIF)

```{r}
# install.packages("car")  # if needed
library(car)
vif(fit_multi)
```

Rule of thumb: VIF much larger than 5–10 suggests strong collinearity (unstable coefficients).

### Heteroscedasticity + robust SE (optional)

```{r}
# install.packages(c("lmtest", "sandwich"))  # if needed
library(lmtest)
library(sandwich)

bptest(fit_multi)  # Breusch–Pagan test

# Robust (HC) standard errors
coeftest(fit_multi, vcov. = vcovHC(fit_multi, type = "HC1"))
```

***

## 3) Regression with factors (dummy coding)

With factors, `lm()` uses a reference level and estimates coefficients as differences relative to that baseline.

```{r}
cars_all <- mtcars |>
  mutate(across(c(cyl, vs, am, gear, carb), as.factor))

fit_factors <- lm(mpg ~ ., data = cars_all)
summary(fit_factors)

# Inspect the design matrix for factor coding
mm <- model.matrix(mpg ~ ., data = cars_all)
dim(mm)
head(mm[, 1:8])
```

Tip: you can change the baseline level via `relevel()` to match the interpretation you want.

***

## 4) Stepwise regression (AIC) + validation

AIC is defined as $$ \mathrm{AIC} = 2k - 2\ln(\hat{L}) $$, i.e., a trade-off between fit and complexity.
Stepwise procedures can be useful for exploration, but they can be unstable and optimistic; good practice is to evaluate performance out-of-sample or via cross-validation.

### Stepwise (AIC) in base R

```{r}
fit_full <- lm(mpg ~ ., data = cars_cont)
fit_null <- lm(mpg ~ 1, data = cars_cont)

fit_step_aic <- step(
  fit_null,
  scope = list(lower = formula(fit_null), upper = formula(fit_full)),
  direction = "both",
  trace = 0
)

summary(fit_step_aic)
AIC(fit_full, fit_step_aic, fit_null)
```

### Out-of-sample check (simple train/test RMSE)

```{r}
set.seed(123)
n <- nrow(cars_cont)
idx_train <- sample.int(n, size = floor(0.8 * n))

train_df <- cars_cont[idx_train, ]
test_df  <- cars_cont[-idx_train, ]

m_full <- lm(mpg ~ ., data = train_df)
m_step <- step(lm(mpg ~ 1, data = train_df),
               scope = list(lower = mpg ~ 1, upper = mpg ~ .),
               direction = "both",
               trace = 0)

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))

pred_full <- predict(m_full, newdata = test_df)
pred_step <- predict(m_step, newdata = test_df)

c(
  RMSE_full = rmse(test_df$mpg, pred_full),
  RMSE_step = rmse(test_df$mpg, pred_step)
)
```

***

## 5) Summary table

```{r, echo=FALSE}
knitr::kable(
  data.frame(
    Topic = c("Simple regression", "Multiple regression", "Factors", "Model selection"),
    Main_function = c("lm()", "lm()", "lm()", "step() + validation"),
    Key_outputs = c(
      "Slope/intercept, CI/PI, residuals",
      "t-tests, R^2, diagnostics, VIF, Cook's D",
      "Reference level, dummy coding, interpretation",
      "AIC trade-off + out-of-sample RMSE/CV"
    ),
    Notes = c(
      "Use PI for individual predictions",
      "Check assumptions; consider robust SE if needed",
      "Relevel factors to control baseline",
      "Stepwise is exploratory; always validate"
    )
  )
)
```

