---
pagetitle: "Lesson 24"
title: "Generalized Linear Models (GLM) in R — Logistic Regression & Model Selection"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Generalized Linear Models (GLM) — Logistic Regression

GLMs generalize linear regression via **link function** $g(\mu) = X\beta$ and **exponential family** density. **Logistic** ($Y \sim Bern(p)$):

$$
\log\frac{p}{1-p} = X\beta, \quad p = \sigma(X\beta) = \frac{1}{1+e^{-X\beta}}
$$

- **Deviance** $D = 2[\ell(\text{saturated}) - \ell(\hat{\mu})]$ measures fit (like $-2\log\lambda$). 
- **AIC** $D + 2p$ balances goodness/complexity. 
- **LRT**: $\text{Deviance}(\text{null}) - \text{Deviance}(\text{alt}) \sim \chi^2_{\Delta df}$.

Lesson: Adult income (>50K?) classification ($n \approx 30k$, 24% positive), feature eng, stepwise, VIF, ROC/PR.

# Step 1: Data import & EDA

```{r}
library(dplyr)
library(stringr)
library(forcats)
library(readr)
library(ggplot2)

data_adult <- read_csv("raw_data/adult.csv",
                       na = c("?", "", " NA"),
                       name_repair = "universal",
                       show_col_types = FALSE) |>
  select(-any_of(c("X", "x"))) |>
  mutate(across(where(is.character), ~forcats::fct_explicit_na(as.factor(.x), "Missing")))

glimpse(data_adult)
prop.table(table(data_adult$income))  # 24% >50K (imbalanced)
```

# Step 2: Numeric EDA & transformation

```{r}
data_adult |>
  select(where(is.numeric)) |>
  summary()

# hours.per.week: cap 99th percentile, log outliers
ggplot(data_adult, aes(hours.per.week)) +
  geom_histogram(bins = 50, fill = "#4AA4DE", alpha = 0.8) +
  labs(title = "Hours/Week Distribution (skewed)")

top1perc <- quantile(data_adult$hours.per.week, 0.99)

data_adult <- data_adult |>
  filter(hours.per.week < top1perc) |> 
  mutate(across(where(is.numeric), ~scale(.x))
  )
```

# Step 3: Categorical recoding (reduce cardinality)

```{r}
data_adult <- data_adult |>
  mutate(
    education = dplyr::case_when(
      education %in% c("Preschool", "1st-4th", "5th-6th") ~ "NoHS",
      education %in% c("7th-8th", "9th", "10th", "11th", "12th") ~ "HSdropout",
      education == "HS-grad" ~ "HS",
      education %in% c("Some-college", "Assoc-acdm", "Assoc-voc") ~ "College",
      education == "Bachelors" ~ "Bachelor",
      TRUE ~ "Advanced"  # Masters, Prof, Doctorate
    ) |> factor(),
    
    marital.status = dplyr::case_when(
      stringr::str_detect(marital.status, "Married") ~ "Married",
      marital.status %in% c("Never-married", "Separated", "Divorced") ~ "Single",
      marital.status == "Widowed" ~ "Widowed"
    ) |> factor()
  )
```

# Step 4: EDA — income by key factors

```{r}
data_adult |>
  dplyr::count(income, gender, sort = TRUE) |>
  mutate(prop = n / sum(n)) |>
  filter(gender == "Male")  # Males 30% >50K vs Females 11%

ggplot(data_adult, aes(x = education, fill = income)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45)) +
  labs(y = "Proportion >50K")
```

# Step 5: Train/valid/test (70/15/15)

```{r}
split_data <- function(df, p_train=0.7, p_valid=0.15, seed=123) {
  set.seed(seed); n <- nrow(df)
  idx_train <- sample(n, floor(p_train*n))
  idx_rem <- setdiff(1:n, idx_train)
  idx_valid <- sample(idx_rem, floor(p_valid*n))
  list(train=df[idx_train,], valid=df[idx_valid,], test=df[-c(idx_train,idx_valid),])
}

spl <- split_data(data_adult)
train_df <- spl$train; valid_df <- spl$valid; test_df <- spl$test
```

# Step 6: Baseline logistic GLM

```{r}
glm_base <- glm(income ~ ., data = train_df, family = binomial())

summary(glm_base)
AIC(glm_base)
```

**Top predictors**: `workclass`, `education`, `marital.status`.

# Step 7: Model diagnostics (VIF, residuals)

```{r}
library(car)
vif(glm_base)  # multicollinearity (<5 OK)

plot(glm_base, which = 1:2)  # residuals vs fitted, Q-Q
```

# Step 8: Stepwise selection (forward/backward)

```{r}
# Both directions
glm_step <- step(glm_base, direction = "both", trace = 0)
summary(glm_step)
anova(glm_base, glm_step, test = "LRT")  # LRT test

AIC(glm_step)  # lower = better fit
```

**LRT p-value**: Improved fit justifies added terms.

# Step 9: Predictions & ROC evaluation

```{r}
library(pROC)
library(caret)

pred_valid_prob <- predict(glm_step, valid_df, type = "response")
pred_valid_class <- ifelse(pred_valid_prob > 0.5, ">50K", "<=50K") |> 
  factor(levels = levels(valid_df$income))

roc_valid <- roc(valid_df$income, pred_valid_prob)
plot(roc_valid, print.auc=TRUE, print.thres="best")

cm_valid <- confusionMatrix(pred_valid_class, valid_df$income, positive = ">50K")
cm_valid$byClass[c("Sensitivity", "Specificity", "Precision", "F1")]
```

**AUC 0.85+**, F1 0.58: Excellent despite imbalance.

Optimal threshold ~0.2 (balance F1).

# Step 10: Final test performance

```{r}
pred_test_prob <- predict(glm_step, test_df, type = "response")
roc_test <- roc(test_df$income, pred_test_prob)
auc_test <- auc(roc_test)

pred_test_class_opt <- ifelse(pred_test_prob > 0.22, ">50K", "<=50K") |> 
  factor(levels = levels(test_df$income))
cm_test <- confusionMatrix(pred_test_class_opt, test_df$income, positive = ">50K")

list(AUC = round(auc_test, 3), F1 = round(cm_test$byClass["F1"], 3))
```

# Comparison: Null vs Stepwise vs Interactions

```{r}
glm_null <- glm(income ~ 1, data = train_df, family = binomial())
anova(glm_null, glm_base, glm_step, test = "LRT")
```

# Warnings & best practices

- **Separation**: Perfect predictors → infinite coefficients (use Firth logistic).  
- **Multicollinearity**: VIF $>10$ → drop/PCA.  
- **Imbalance**: ROC/PR, threshold $\neq 0.5$, SMOTE.  
- **Linearity logit**: Check via loess/BOX-TIDY.  
- **AIC/LRT**: Relative fit; validate on hold-out.

# Summary

You learned GLM logistic con `glm()` to:

- Model **log-odds**: $\log(p/(1-p)) = X\beta$.  
- Select via **stepwise + LRT**: $\Delta D \sim \chi^2_{\Delta df}$.  
- Diagnose VIF/residuals, evaluate **ROC/PR/F1**.  
- Optimize **threshold** for imbalance.  
