---
pagetitle: "Lesson 29"
title: "Support Vector Machine in R (caret + kernlab) — Titanic Classification"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Support Vector Machine (SVM) on Titanic

Support Vector Machines are powerful classifiers that try to find a decision boundary with maximum margin.
In this lesson we use an SVM with **Radial Basis Function (RBF)** kernel.

In this lesson we will:
- Import Titanic data from a local CSV
- Clean the dataset (drop columns + fix types)
- One-hot encode categorical variables (SVM needs numeric predictors)
- Create train/test split
- Train an SVM (RBF) with cross-validation
- Evaluate accuracy + confusion matrix
- Tune `sigma` and `C`

# Step 1: Import the data

```{r}
library(dplyr)

path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

dim(titanic)
head(titanic, 3)
```

# Step 2: Clean and prepare data

```{r}
titanic_clean <- titanic |>
  select(-any_of(c("home.dest", "cabin", "name", "X", "x", "ticket"))) |>
  filter(embarked != "?") |>
  mutate(
    pclass = factor(
      pclass,
      levels = c(1, 2, 3),
      labels = c("Upper", "Middle", "Lower")
    ),
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = as.numeric(fare),
    sibsp = as.numeric(sibsp),
    parch = as.numeric(parch)
  ) |>
  na.omit()

glimpse(titanic_clean)
table(titanic_clean$survived)
```

## Encoding (SVM needs numeric predictors)

We use one-hot encoding with `model.matrix()`.

```{r}
X_all <- model.matrix(survived ~ . - 1, data = titanic_clean)
y_all <- titanic_clean$survived

dim(X_all)
table(y_all)
```

# Step 3: Train/test split

```{r}
set.seed(123)

n <- nrow(X_all)
idx_train <- sample.int(n, size = floor(0.8 * n))

X_train <- X_all[idx_train, , drop = FALSE]
y_train <- y_all[idx_train]

X_test <- X_all[-idx_train, , drop = FALSE]
y_test <- y_all[-idx_train]

c(n_train = nrow(X_train), n_test = nrow(X_test))
```

# Step 4: Train a baseline SVM (RBF) with cross-validation

Notes:
- SVM is sensitive to feature scaling, so we apply `center` and `scale`.
- `svmRadial` in caret tunes `sigma` and `C` (cost).

```{r}
# install.packages(c("caret", "kernlab", "e1071"))
library(caret)
library(kernlab)
library(e1071)

set.seed(123)

ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE
)

svm_base <- train(
  x = X_train,
  y = y_train,
  method = "svmRadial",
  metric = "Accuracy",
  preProcess = c("center", "scale"),
  trControl = ctrl,
  tuneLength = 6
)

svm_base
svm_base$bestTune
```

# Step 5: Predictions

```{r}
pred_class <- predict(svm_base, newdata = X_test)
head(pred_class)
```

# Step 6: Evaluation (confusion matrix + accuracy)

```{r}
confusionMatrix(pred_class, y_test)
```

# Step 7: Manual tuning grid (sigma, C)

A small explicit grid search (you can expand it if needed).

```{r}
grid <- expand.grid(
  sigma = c(0.001, 0.01, 0.05, 0.1),
  C     = c(0.25, 0.5, 1, 2, 4)
)

set.seed(123)
svm_tuned <- train(
  x = X_train,
  y = y_train,
  method = "svmRadial",
  metric = "Accuracy",
  preProcess = c("center", "scale"),
  trControl = ctrl,
  tuneGrid = grid
)

svm_tuned
svm_tuned$bestTune
```

Evaluate tuned model:

```{r}
pred_class2 <- predict(svm_tuned, newdata = X_test)
confusionMatrix(pred_class2, y_test)
```

# Summary

You learned how to:
- clean Titanic data and create numeric predictors via one-hot encoding
- train an SVM with RBF kernel using cross-validation
- scale features (center/scale)
- tune `sigma` and `C` and evaluate with a confusion matrix
