---
pagetitle: "Lesson 6"
title: "List in R: Create, Select Elements with Example"
# author: "Dott. Gianluca Sottile"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

<br>

# What is a List?
***
A list is a great tool to store many kinds of object in the order expected. We can include matrices, vectors data frames or lists. We can imagine a list as a bag in which we want to put many different items. When we need to use an item, we open the bag and use it. A list is similar; we can store a collection of objects and use them when we need them.

# How to Create a List
***
We can use list() function to create a list.

```{r, don-eval, eval=FALSE}
list(element_1, ...)
```

__Arguments:__

+ element_1: store any type of R object
+ ...: pass as many objects as specifying. each object needs to be separated by a comma

In the example below, we create three different objects, a vector, a matrix and a data frame.

__Step 1)__ Create a Vector

```{r}
# Vector with numeric from 1 up to 5
vect  <- 1:5
```

__Step 2)__ Create a Matrices

```{r}
# A 2x 5 matrix
mat  <- matrix(1:9, ncol = 5)
dim(mat)
```

__Step 3)__ Create Data Frame

```{r}
# select the 10th row of the built-in R data set EuStockMarkets
df <- EuStockMarkets[1:10,]
```

__Step 4)__ Create a List

Now, we can put the three object into a list.

```{r}
# Construct list with these vec, mat, and df:
my_list <- list(vect, mat, df)
my_list
```

# Select Elements from List
***
After we built our list, we can access it quite easily. We need to use the [[index]] to select an element in a list. The value inside the double square bracket represents the position of the item in a list we want to extract. For instance, we pass 2 inside the parenthesis, R returns the second element listed.

Let's try to select the second items of the list named my_list, we use `my_list[[2]]`

```{r}
# Print second element of the list
my_list[[2]]
```

# Built-in Data Frame
***
Before to create our own data frame, we can have a look at the R data set available online. The prison dataset is a 714x5 dimension. We can get a quick look at the bottom of the data frame with `tail()` function. By analogy, `head()` displays the top of the data frame. You can specify the number of rows shown with `head(df, 5)`. We will learn more about the function `read.csv()` in future tutorial.

```{r}
PATH <-'https://raw.githubusercontent.com/guru99-edu/R-Programming/master/prison.csv'
df <- read.csv(PATH)[1:5]
head(df, 5)
```

We can check the structure of the data frame with str:

```{r}
# Structure of the data
str(df)
```

All variables are stored in the __numerical__ format.

