---
pagetitle: "Lesson 28"
title: "Gradient Boosting in R (XGBoost) — Titanic Classification"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Gradient Boosting (XGBoost) on Titanic

Gradient boosting builds trees sequentially: each new tree focuses on correcting the mistakes of the previous ensemble.
XGBoost is a highly optimized implementation of gradient boosting with practical features like regularization and early stopping.

In this lesson we will:
- Import Titanic data from a URL
- Clean and encode features for XGBoost
- Create train/test split
- Train XGBoost for binary classification
- Evaluate accuracy + confusion matrix
- Use cross-validation with early stopping
- Inspect feature importance

# Step 1: Import the data

```{r}
library(dplyr)

path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

dim(titanic)
head(titanic, 3)
```

# Step 2: Clean and prepare data

We follow a simple approach similar to the Decision Tree lesson:
- Drop columns that are not useful or require heavy feature engineering (IDs, names, ticket, cabin, etc.)
- Convert categorical variables to factors
- Remove missing values (for simplicity)

```{r}
titanic_clean <- titanic |>
  select(-c(home.dest, cabin, name, x, ticket)) |>
  filter(embarked != "?") |>
  mutate(
    pclass = factor(
      pclass,
      levels = c(1, 2, 3),
      labels = c("Upper", "Middle", "Lower")
    ),
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = as.numeric(fare)
  ) |>
  na.omit()

glimpse(titanic_clean)
```

## Why we need encoding

XGBoost expects a numeric matrix.
A common and reliable approach is one-hot encoding via `model.matrix()`.

```{r}
# Create design matrix (one-hot encoding for factors)
X_all <- model.matrix(survived ~ . - 1, data = titanic_clean)

# Label as 0/1 (XGBoost wants numeric labels for binary logistic)
y_all <- ifelse(titanic_clean$survived == "Yes", 1, 0)

dim(X_all)
table(y_all)
```

# Step 3: Train/test split

```{r}
set.seed(123)

n <- nrow(X_all)
idx_train <- sample.int(n, size = floor(0.8 * n))

X_train <- X_all[idx_train, , drop = FALSE]
y_train <- y_all[idx_train]

X_test <- X_all[-idx_train, , drop = FALSE]
y_test <- y_all[-idx_train]

c(n_train = nrow(X_train), n_test = nrow(X_test))
```

# Step 4: Train a baseline XGBoost model

We use binary logistic objective:
- output is a probability in [0, 1]
- we then choose a threshold (e.g., 0.5) for class prediction

```{r}
library(xgboost)

dtrain <- xgb.DMatrix(data = X_train, label = y_train)
dtest  <- xgb.DMatrix(data = X_test,  label = y_test)

params <- list(
  objective = "binary:logistic",
  eval_metric = "logloss",
  max_depth = 3,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8
)

set.seed(123)
xgb_fit <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  watchlist = list(train = dtrain, test = dtest),
  verbose = 0
)

xgb_fit
```

# Step 5: Predictions and evaluation

```{r}
pred_prob <- predict(xgb_fit, newdata = dtest)
pred_class <- ifelse(pred_prob >= 0.5, 1, 0)

conf_mat <- table(actual = y_test, predicted = pred_class)
conf_mat

accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
accuracy
```

# Step 6: Cross-validation + early stopping (choose nrounds)

Early stopping finds the number of boosting rounds that minimizes the validation loss and stops when there is no improvement for some rounds.

```{r}
set.seed(123)

cv <- xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 1000,
  nfold = 5,
  early_stopping_rounds = 25,
  verbose = 0
)

cv$early_stop$best_iteration
```

Train a final model using the best number of rounds:

```{r}
best_nrounds <- cv$early_stop$best_iteration

set.seed(123)
xgb_final <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  watchlist = list(train = dtrain, test = dtest),
  verbose = 0
)

pred_prob2 <- predict(xgb_final, newdata = dtest)
pred_class2 <- ifelse(pred_prob2 >= 0.5, 1, 0)

conf_mat2 <- table(actual = y_test, predicted = pred_class2)
conf_mat2

accuracy2 <- sum(diag(conf_mat2)) / sum(conf_mat2)
accuracy2
```

# Step 7: Feature importance

```{r}
imp <- xgb.importance(model = xgb_final, feature_names = colnames(X_train))
head(imp, 10)

xgb.plot.importance(imp, top_n = 15)
```

# (Optional) Step 8: Simple tuning grid

This is a small grid search to illustrate the idea (not exhaustive).

```{r}
grid <- expand.grid(
  max_depth = c(2, 3, 4),
  eta = c(0.05, 0.1),
  subsample = c(0.8, 1.0),
  colsample_bytree = c(0.8, 1.0)
)

grid$best_iter <- NA_integer_
grid$best_auc <- NA_real_

set.seed(123)

for (i in seq_len(nrow(grid))) {
  params_i <- list(
    objective = "binary:logistic",
    eval_metric = "auc",
    max_depth = grid$max_depth[i],
    eta = grid$eta[i],
    subsample = grid$subsample[i],
    colsample_bytree = grid$colsample_bytree[i]
  )

  cv_i <- xgb.cv(
    params = params_i,
    data = dtrain,
    nrounds = 800,
    nfold = 5,
    early_stopping_rounds = 25,
    verbose = 0
  )

  grid$best_iter[i] <- cv_i$early_stop$best_iteration
  grid$best_auc[i] <- cv_i$evaluation_log$train_auc_mean[cv_i$early_stop$best_iteration]
}

grid <- grid[order(grid$best_auc, decreasing = TRUE), ]
head(grid, 10)
```

# Summary

You learned how to:
- clean and encode Titanic data for XGBoost
- train a gradient boosting model for binary classification
- evaluate with a confusion matrix and accuracy
- use cross-validation + early stopping to pick the number of trees
- inspect feature importance
