---
pagetitle: "Lesson 2"
title: "R Data Types, Variables, and Operators (with Examples)"
# author: "Gianluca Sottile"
editor_options:
  chunk_output_type: inline
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# Basic data types {#buttons .tabset .tabset-fade .tabset-pills}

R works with several data types and data structures. The most common ones are:

- Scalars (single values)
- Vectors (numeric, character, logical)
- Matrices
- Data frames
- Lists

**Basic atomic types**

- `4.5` is a decimal number (type: **numeric**).
- `4L` is an integer (type: **integer**). Integers are a subset of numeric values.
- `TRUE` / `FALSE` are Boolean values (type: **logical**).
- Text in quotes (`" "` or `' '`) is a string (type: **character**).

You can check the type of an object with `class()`.

## Example
```{r}
# Numeric (stored as "double" by default)
x <- 28
class(x)

# Character (string)
y <- "R is fantastic"
class(y)

# Logical
z <- TRUE
class(z)

# Integer (note the L suffix)
w <- 4L
class(w)
```

# Variables {#buttons1 .tabset .tabset-fade .tabset-pills}

Variables store values and are fundamental in programming. In R, a variable can store almost anything: a number, a vector, a dataset, a model, or the result of a computation.

To create a variable, assign a value to a name. Variable names should not contain spaces; use `_` if you want to separate words.

In R, assignment is typically done with `<-` (you may also see `=` in some contexts).

## Syntax
```{r, eval=FALSE}
name_of_variable <- value
# or
name_of_variable = value
```

## Examples
```{r}
# Create variables
x <- 42
y <- 10

# Print variables
x
y

# Use variables in an expression
x - y
```

# Vectors {#buttons2 .tabset .tabset-fade .tabset-pills}

A vector is a one-dimensional collection of values. A vector usually contains values of the same type (all numeric, all character, etc.). You can create vectors with `c()`.

## Examples
```{r}
# Numeric vector
vec_num <- c(1, 10, 49)
vec_num

# Character vector
vec_chr <- c("a", "b", "c")
vec_chr

# Logical vector
vec_lgl <- c(TRUE, FALSE, TRUE)
vec_lgl
```

You can perform arithmetic on vectors element-by-element (as long as lengths are compatible).

```{r}
vect_1 <- c(1, 3, 5)
vect_2 <- c(2, 4, 6)

sum_vect <- vect_1 + vect_2
sum_vect
```

In R, slicing uses square brackets `[]`. For example, `x[1:5]` selects elements from position 1 to 5.

```{r}
slice_vector <- 1:10
slice_vector[1:5]
```

The shortest way to create a sequence is the `:` operator.

```{r}
1:10
```

# Arithmetic operators {#buttons3 .tabset .tabset-fade .tabset-pills}

R supports the most common arithmetic operators:

```{r}
ops_arith <- data.frame(
  Operator = c("+", "-", "*", "/", "^", "%%"),
  Description = c(
    "Addition",
    "Subtraction",
    "Multiplication",
    "Division",
    "Exponentiation",
    "Modulo (remainder)"
  )
)

knitr::kable(ops_arith)
```

Notes:

- Lines starting with `#` are comments and are ignored by R.
- When you run code, the result is printed as output (in rendered documents you often see output prefixed by `##`).

## Examples
```{r}
# Addition
3 + 4

# Multiplication
3 * 5

# Division (use parentheses to control precedence)
(5 + 5) / 2

# Exponentiation
2^5

# Modulo (remainder)
28%% 6
```

# Logical operators {#buttons4 .tabset .tabset-fade .tabset-pills}

Logical operators return `TRUE` or `FALSE`. They are essential for filtering data and writing conditions.

```{r}
ops_logical <- data.frame(
  Operator = c("<", "<=", ">", ">=", "==", "!=", "!", "&", "|"),
  Description = c(
    "Less than",
    "Less than or equal to",
    "Greater than",
    "Greater than or equal to",
    "Equal to",
    "Not equal to",
    "NOT",
    "AND (vectorised)",
    "OR (vectorised)"
  )
)

knitr::kable(ops_logical)
```

To select elements that satisfy a condition, use:

```{r, eval=FALSE}
x[condition]
```

## Example 1: Test a condition
```{r}
logical_vector <- 1:10
logical_vector > 5
```

## Example 2: Filter values using a condition
```{r}
logical_vector[logical_vector > 5]
```

## Example 3: Combine conditions
```{r}
logical_vector[(logical_vector > 4) & (logical_vector < 7)]
```
