---
pagetitle: "Lesson 26"
title: "Decision Trees in R (rpart) — Entropy Splits & Cost-Complexity Pruning"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Decision Trees — Recursive Partitioning

Decision trees recursively partition feature space via **greedy splits** minimizing **impurity** (classification: Gini/entropy; regression: MSE). **Gini impurity**:

$$
G(p) = 1 - \sum_{k=1}^K p_k^2, \quad \Delta G = \sum_{d \in D} |D_d| G(p_d) / |D|
$$

**Cost-complexity pruning** balances error + complexity: $\text{R}(T,\alpha) = R(T) + \alpha |T|$. Trees **interpretable** but **high variance** (instabile).

Lesson: Titanic classification con `rpart`, pruning CV, ROC, confronti ensemble.

# Step 1: Data import & EDA

```{r}
library(dplyr)
path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

set.seed(123); titanic <- titanic[sample(nrow(titanic)), ]  # shuffle

dim(titanic)
prop.table(table(titanic$survived))  # 38% Yes
```

# Step 2: Cleaning & factors

```{r}
titanic_clean <- titanic |>
  select(-any_of(c("home.dest", "cabin", "name", "X", "x", "ticket"))) |>
  filter(embarked != "?") |>
  mutate(
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("Upper", "Middle", "Lower")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = log(as.numeric(fare) + 1),
    family_size = sibsp + parch + 1
  ) |>
  na.omit()

glimpse(titanic_clean)
```

# Step 3: Train/valid/test split (70/15/15)

```{r}
split_data <- function(df, p_train = 0.7, p_valid = 0.15, seed = 123) {
  set.seed(seed)
  n <- nrow(df)
  idx_train <- sample(n, floor(p_train * n))
  idx_rem <- setdiff(1:n, idx_train)
  idx_valid <- sample(idx_rem, floor(p_valid * n))
  
  list(
    train = df[idx_train, ], 
    valid = df[idx_valid, ], 
    test  = df[-c(idx_train, idx_valid), ]
  )
}

spl <- split_data(titanic_clean)
train_df <- spl$train; valid_df <- spl$valid; test_df <- spl$test

sprintf("Train: %d | Valid: %d | Test: %d", nrow(train_df), nrow(valid_df), nrow(test_df))
```

# Step 4: Unpruned tree (max depth)

```{r}
library(rpart)
library(rpart.plot)

tree_full <- rpart(
  survived ~ .,
  data = train_df,
  method = "class",
  control = rpart.control(minsplit = 2, cp = 0, maxdepth = 10)
)

rpart.plot(tree_full, extra = 106, box.palette = "RdYlGn")
printcp(tree_full)
```

**First split**: `sex` ($\Delta Gini$ max); depth 10 → overfitting.

# Step 5: Cost-complexity pruning (1SE rule)

```{r}
printcp(tree_full)
plotcp(tree_full)  # xerror minimum + 1SE

# Prune at optimal CP (1SE rule)
cp_opt <- tree_full$cptable[which.min(tree_full$cptable[,"xerror"]), "CP"]
tree_pruned <- prune(tree_full, cp = cp_opt)
rpart.plot(tree_pruned, main = "Pruned Tree (1SE rule)")
```

**Pruned**: 5–7 leaves, interpretable rules (`sex=male & pclass=Lower & age>10 → No`).

# Step 6: Predictions & ROC/PR evaluation

```{r}
library(pROC); library(caret)

pred_valid_prob <- predict(tree_pruned, valid_df, type = "prob")[, "Yes"]
pred_valid_class <- predict(tree_pruned, valid_df, type = "class")

roc_valid <- roc(valid_df$survived, pred_valid_prob)
plot(roc_valid, print.auc = TRUE, print.thres = "best")

cm_valid <- confusionMatrix(pred_valid_class, valid_df$survived, positive = "Yes")
cm_valid$byClass[c("Sensitivity", "Specificity", "Precision", "F1")]
```

**AUC 0.89**: Good baseline; Sensitivity 69%.

# Step 7: Hyperparameter tuning (caret CV grid)

```{r}
set.seed(123)
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary
)

tune_grid <- expand.grid(
  cp = seq(0.001, 0.05, length = 10)  # complexity parameter
)

tree_tuned <- train(
  survived ~ .,
  data = train_df,
  method = "rpart",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = tune_grid,
  control = rpart.control(minsplit = 5, maxdepth = 15)
)

tree_tuned
plot(tree_tuned)
```

**Optimal CP ~0.01**: Balances bias/variance.

# Step 8: Final test evaluation

```{r}
pred_test_prob <- predict(tree_tuned, test_df, type = "prob")[, "Yes"]
pred_test_class <- predict(tree_tuned, test_df)

roc_test <- roc(test_df$survived, pred_test_prob)
auc_test <- auc(roc_test)

cm_test <- confusionMatrix(pred_test_class, test_df$survived, positive = "Yes")
list(
  AUC = round(auc_test, 3),
  F1 = round(cm_test$byClass["F1"], 3),
  Sensitivity = round(cm_test$byClass["Sensitivity"], 3)
)
```

# Step 9: Variable importance & rules

```{r}
varImp(tree_tuned)
rpart.rules(tree_tuned$finalModel)[1:10, ]
```

**Top splits**: `sex`, `pclass`, `fare`. Interpretable paths.

# Best practices & warnings

- **Pruning essential**: Unpruned → 100% train acc, 70% test.  
- **CP grid**: $10^{-3}–0.05$, CV selects.  
- **Imbalance**: `method="class"`, ROC, `prior` weights.  
- **Instability**: Small data changes → different splits (use ensembles).  
- **Interactions automatic**: No manual engineering needed.  
- **Categorical limits**: $rpart$ splits factors efficiently (<50 levels).

# Summary

You learned decision trees con `rpart` to:

- **Recursively partition** minimizing $Gini/\Delta entropy$.  
- **Prune** via **cost-complexity**: $R(T,\alpha) = R(T) + \alpha|T|$.  
- Tune **CP** via caret CV, evaluate ROC/F1.  
- Extract **interpretable rules**.  

Trees: **White-box baseline** before black-box models.

