---
pagetitle: "Lesson 2"
title: "R Data Types, Arithmetic & Logical Operators with Example"
# author: "Dott. Gianluca Sottile"
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

<br>

# Basic data types {#buttons .tabset .tabset-fade .tabset-pills}
***
R Programming works with numerous data types, including

+ Scalars
+ Vectors (numerical, character, logical)
+ Matrices
+ Data frames
+ Lists

__Basics types__

+ 4.5 is a decimal value called numerics.
+ 4 is a natural value called integers. Integers are also numerics.
+ TRUE or FALSE is a Boolean value called logical.
+ The value inside " " or ' ' are text (string). They are called characters.

We can check the type of a variable with the class function

## Example 1:
```{r}
# Declare variables of different types
# Numeric
x <- 28
class(x)
```

## Example 2:
```{r}
# String
y <- "R is Fantastic"
class(y)
```

## Example 3:
```{r}
# Boolean
z <- TRUE
class(z)
```

# Variables {#buttons1 .tabset .tabset-fade .tabset-pills}
***
Variables store values and are an important component in programming, especially for a data scientist. A variable can store a number, an object, a statistical result, vector, dataset, a model prediction basically anything R outputs. We can use that variable later simply by calling the name of the variable.

To declare a variable, we need to assign a variable name. The name should not have space. We can use _ to connect to words.

To add a value to the variable, use <- or =.

Here is the syntax:

```{r, don-eval, eval=FALSE}
# First way to declare a variable:  use the `<-`
name_of_variable <- value
# Second way to declare a variable:  use the `=`
name_of_variable = value
```

In the command line, we can write the following codes to see what happens:

## Example 1:
```{r}
# Print variable x
x <- 42
```

## Example 2:
```{r}
y  <- 10
y
```

## Example 3:
```{r}
# We call x and y and apply a subtraction
x - y
```

# Vectors {#buttons2 .tabset .tabset-fade .tabset-pills}
***
A vector is a one-dimensional array. We can create a vector with all the basic data type we learnt before. The simplest way to build a vector in R, is to use the c command.

## Example 1:
```{r}
# Numerical
vec_num <- c(1, 10, 49)
vec_num
```

## Example 2:
```{r}
# Character 
vec_chr <- c("a", "b", "c")
vec_chr
```

## Example 3:
```{r}
# Boolean 
vec_bool <-  c(TRUE, FALSE, TRUE)
vec_bool
```

## Example 4:
We can do arithmetic calculations on vectors.
```{r}
# Create the vectors
vect_1 <- c(1, 3, 5)
vect_2 <- c(2, 4, 6)
# Take the sum of A_vector and B_vector
sum_vect <- vect_1 + vect_2
# Print out total_vector
sum_vect
```

## Example 5:
In R, it is possible to slice a vector. In some occasion, we are interested in only the first five rows of a vector. We can use the `[1:5]` command to extract the value 1 to 5.
```{r}
# Slice the first five rows of the vector
slice_vector <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
slice_vector[1:5]
```

## Example 6:
The shortest way to create a range of value is to use the: between two numbers. For instance, from the above example, we can write `c(1:10)` to create a vector of value from one to ten.
```{r}
# Faster way to create adjacent values
c(1:10)
```

# Arithmetic Operators {#buttons3 .tabset .tabset-fade .tabset-pills}
***
We will first see the basic arithmetic operations in R. The following operators stand for:

```{r, echo=FALSE}
library(knitr)
library(kableExtra)
dt <- data.frame(
  "Operator" = c("+", "-", "*", "/", "^ or **"), 
  "Description" = c("Addition", "Subtraction", "Multiplication", "Division", "Exponentiation"))
kable(dt) %>% kable_styling(bootstrap_options = c("striped", "hover"))
```

<!-- | Operator | Description |  -->
<!-- |:---------|:------------| -->
<!-- | + |	Addition | -->
<!-- | - |	Subtraction | -->
<!-- | * |	Multiplication | -->
<!-- | / |	Division | -->
<!-- | ^ or ** |	Exponentiation | -->

You can easily copy and paste the above R code into Rstudio Console. The __output__ is displayed after the character #. For instance, we write the code print('Hello world') the output will be ##[1] "Hello world".

The ## means we print an output and the number in the square bracket ([1]) is the number of the display

The sentences starting with # __annotation__. We can use # inside an R script to add any comment we want. R won't read it during the running time.

## Example 1:
```{r}
# An addition
3 + 4
```

## Example 2:
```{r}
# A multiplication
3 * 5
```

## Example 3:
```{r}
# A division
(5 + 5) / 2
```

## Example 4:
```{r}
# Exponentiation
2^5
```

## Example 5:
```{r}
# Modulo
28 %% 6
```

# Logical Operators {#buttons4 .tabset .tabset-fade .tabset-pills}
***
With logical operators, we want to return values inside the vector based on logical conditions. Following is a detailed list of logical operators available in R

```{r, echo=FALSE}
library(knitr)
library(kableExtra)
dt <- data.frame(
  "Operator" = c("<", "<=", ">", ">=", "==", "!=", "!x", "x & y", "isTRUE(x"), 
  "Description" = c("Less than", "Less than or equal to", "Greater than", "Greater than or equal to", "Exactly equal to", "Not equal to", "Not x", "x AND y", "Test if x is TRUE"))
kable(dt) %>% kable_styling(bootstrap_options = c("striped", "hover"))
```

<!-- | Operator | Description |  -->
<!-- |:---------|:------------| -->
<!-- | < |	Less than | -->
<!-- | <= | Less than or equal to | -->
<!-- | > |	Greater than | -->
<!-- | >= | Greater than or equal to | -->
<!-- | == | Exactly equal to | -->
<!-- | != | Not equal to | -->
<!-- | !x | Not X | -->
<!-- | x & y |	x AND y | -->
<!-- | isTRUE(x) |	Test if x is TRUE | -->

The logical statements in R are wrapped inside the `[ ]`. We can add many conditional statements as we like but we need to include them in a parenthesis. We can follow this structure to create a conditional statement:

```{r, don-eval1, eval=FALSE}
variable_name[(conditional_statement)]
```

With variable_name referring to the variable, we want to use for the statement. We create the logical statement i.e. `variable_name > 0`. Finally, we use the square bracket to finalize the logical statement. Below, an example of a logical statement.

## Example 1:
```{r}
# Create a vector from 1 to 10
logical_vector <- c(1:10)
logical_vector > 5
```

In the output above, R reads each value and compares it to the statement `logical_vector > 5`. If the value is strictly superior to five, then the condition is TRUE, otherwise FALSE. R returns a vector of TRUE and FALSE.

## Example 2:
In the example below, we want to extract the values that only meet the condition 'is strictly superior to five'. For that, we can wrap the condition inside a square bracket precede by the vector containing the values.
```{r}
# Print value strictly above 5
logical_vector[(logical_vector > 5)]
```

## Example 3:
```{r}
# Print 5 and 6
logical_vector <- c(1:10)
logical_vector[(logical_vector > 4) & (logical_vector < 7)]
```


