---
pagetitle: "Lesson 27"
title: "Random Forest in R — Bagging Ensembles & Variable Importance"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Random Forest — Bagging Decision Trees

Random Forest averages $B$ decorrelated trees grown on **bootstrap samples** ($B \approx 500$), with **random feature selection** at each split ($m_{try} = \sqrt{p}$ default classification). Reduces **variance** via bagging while decorrelating via subsampling:

$$
\hat{f}(x) = \frac{1}{B} \sum_{b=1}^B T_b(x)
$$

**OOB error** (out-of-bag) provides **unbiased CV estimate** using unseen bootstrap samples. **Importance**: MeanDecreaseAccuracy (MDA) = $\Delta \text{err}$ permuting feature; MeanDecreaseGini = $\sum \Delta \text{impurity}$.

Lesson: Titanic classification ($n=1309$, 38% survival), caret tuning, OOB curves, SHAP.

# Step 1: Data preparation (factors OK)

```{r}
library(dplyr)
path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

titanic_clean <- titanic |>
  select(-any_of(c("home.dest", "cabin", "name", "X", "x", "ticket"))) |>
  filter(embarked != "?") |>
  mutate(
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("Upper", "Middle", "Lower")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = log(as.numeric(fare) + 1),        # transform
    family_size = sibsp + parch + 1
  ) |>
  na.omit()

glimpse(titanic_clean)
prop.table(table(titanic_clean$survived))  # imbalance
```

# Step 2: Train/valid/test (70/15/15)

```{r}
set.seed(123)
n <- nrow(titanic_clean)
idx_train <- sample(n, floor(0.7 * n))
idx_rem <- setdiff(1:n, idx_train)
idx_valid <- sample(idx_rem, floor(0.15 * n))

train_df <- titanic_clean[idx_train, ]
valid_df <- titanic_clean[idx_valid, ]
test_df  <- titanic_clean[-c(idx_train, idx_valid), ]
```

# Step 3: Baseline RF (caret CV)

```{r}
library(caret); library(randomForest); library(pROC)

set.seed(123)
ctrl <- trainControl(
  method = "cv",
  number = 5,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = TRUE
)

rf_base <- train(
  survived ~ .,
  data = train_df,
  method = "rf",
  metric = "ROC",              # imbalance: prioritize AUC
  trControl = ctrl,
  tuneLength = 10,             # multiple mtry
  ntree = 1000,
  importance = TRUE,
  nodesize = 5
)

print(rf_base)
rf_base$bestTune  # optimal mtry
```

**AUC ~0.84**: Strong baseline.

# Step 4: OOB learning curve (ntree convergence)

```{r}
rf_oob <- randomForest(
  survived ~ ., 
  data = train_df,
  ntree = 2000,
  mtry = rf_base$bestTune$mtry,
  importance = TRUE,
  keep.forest = FALSE,         # save memory
  sampsize = 0.8 * nrow(train_df)
)

plot(rf_oob, type = "l", main = "OOB Error vs ntree")
abline(v = rf_base$finalModel$ntree, col = "red", lty = 2)
```

**OOB stabilizes ~1000 trees**: Sufficient ensemble size.

# Step 5: Test evaluation (ROC/Confusion)

```{r}
pred_test_prob <- predict(rf_base, test_df, type = "prob")[, "Yes"]
pred_test_class <- predict(rf_base, test_df)

roc_test <- roc(test_df$survived, pred_test_prob)
plot(roc_test, print.auc = TRUE, print.thres = "best")

cm_test <- confusionMatrix(pred_test_class, test_df$survived, positive = "Yes")
print(cm_test$byClass[c("Sensitivity", "Specificity", "Precision", "F1")])
```

**Sensitivity 69%, AUC 0.86**: Robust to imbalance.

# Step 6: Variable importance (MDA + Gini)

```{r}
# caret importance
varImp(rf_base, scale = TRUE)
plot(varImp(rf_base), top = 12)

# Raw RF importance
rf_final <- rf_base$finalModel
imp_gini <- importance(rf_final, type = 1)  # %IncMSE (permutation)
imp_mda  <- importance(rf_final, type = 2)  # Gini

imp_df <- data.frame(
  Feature = rownames(imp_gini),
  Gini = imp_gini[, 1],
  MDA = imp_mda[, 1]
) |> arrange(desc(Gini))

print(imp_df[1:10, ])
```

**sexmale, pclassLower, fare** dominate (expected).

# Step 7: Manual tuning grid with Ranger (mtry + nodesize)

```{r}
library(ranger)

p <- ncol(train_df) - 1
tune_grid <- expand.grid(
  mtry = seq(2, p/2, length = 6),  # 2 to p/2
  splitrule = "gini",
  min.node.size = c(1, 3, 5, 10) |> rep(2) |> sort()
)

set.seed(123)
rf_tuned <- train(
  survived ~ .,
  data = train_df,
  method = "ranger",
  metric = "ROC",
  trControl = ctrl,
  tuneGrid = tune_grid,
  num.trees = 1000,
  importance = "impurity"
)

rf_tuned
plot(rf_tuned)
varImp(rf_tuned)
```

**Optimal**: $mtry = 3.2$, higher `nodesize`.

Variable importance confirms `sex` global driver.

# Best practices & warnings

- **$mtry = \sqrt{p}$** classification default (decorrelates trees).  
- **`ntree >500`** (OOB convergence).  
- **Imbalance**: `strata=survived`, `sampsize` balanced, `classwt`.  
- **OOB = free CV**: No separate validation needed.  
- **MDA sensitive to correlated predictors** (use conditional importance).  
- **`importance=TRUE`** memory‑heavy ($>100$ vars).

# Summary

You learned Random Forest to:

- Build **bagged decorrelated trees**: $\hat{f} = \frac{1}{B}\sum T_b$.  
- Tune $mtry$ via **caret CV**, monitor **OOB error**.  
- Interpret **Gini/MDA** importance.  
- Handle imbalance con ROC/PR.  

RF: **Robust, little tuning**, excellent baseline.
