---
pagetitle: "Lesson 28"
title: "Gradient Boosting (XGBoost) in R — Functional Gradient Descent & Tuning"
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# Gradient Boosting Machines (GBM) — XGBoost

Gradient Boosting builds an ensemble $F_M(x) = \sum_{m=1}^M T_m(x)$ by **functional gradient descent**: each tree $T_m$ fits **pseudo-residuals** $r_{im} = -[\mathcal{L}(y_i, F_{m-1}(x_i))]$ of the **base learner loss** $\mathcal{L}$ (logistic $logloss$ here), with **regularization**:

$$
\mathcal{L} = - \sum [y_i \log p_i + (1-y_i)\log(1-p_i)] + \Omega(F)
$$

XGBoost adds **L1/L2 tree regularization**, **subsampling**, **early stopping**. Extremely efficient (histogram binning, sparsity-aware).

Lesson: Complete pipeline on Titanic ($n=1309$, 38% survival), CV tuning, SHAP importance.

# Step 1: Data import & EDA

```{r}
library(dplyr)
path <- "raw_data/titanic_data.csv"
titanic <- read.csv(path, stringsAsFactors = FALSE)

dim(titanic)
prop.table(table(titanic$survived))  # 38% Yes (imbalanced)
head(titanic)
```

# Step 2: Feature engineering & encoding

```{r}
titanic_clean <- titanic |>
  select(-any_of(c("home.dest", "cabin", "name", "X", "x", "ticket"))) |>
  filter(embarked != "?") |>
  mutate(
    survived = factor(survived, levels = c(0, 1), labels = c("No", "Yes")),
    pclass = factor(pclass, levels = c(1, 2, 3), labels = c("Upper", "Middle", "Lower")),
    sex = factor(sex),
    embarked = factor(embarked),
    age = as.numeric(age),
    fare = as.numeric(fare),
    sibsp = as.numeric(sibsp),
    parch = as.numeric(parch),
    family_size = sibsp + parch + 1,
    age_child = ifelse(age < 18, 1, 0),  # interaction
    fare_log = log(fare + 1)
  ) |>
  na.omit()

glimpse(titanic_clean)
```

**XGBoost DMatrix** (numeric + labels 0/1):
```{r}
X_all <- model.matrix(survived ~ . - 1, data = titanic_clean)
y_all <- as.numeric(titanic_clean$survived == "Yes")  # 0/1 numeric

dim(X_all)
```

# Step 3: Train/valid/test split (70/15/15)

```{r}
library(xgboost)

set.seed(123)
n <- nrow(X_all)
idx_train <- sample(n, floor(0.7 * n))
idx_rem <- setdiff(1:n, idx_train)
idx_valid <- sample(idx_rem, floor(0.15 * n))

X_train <- X_all[idx_train, ]; y_train <- y_all[idx_train]
X_valid <- X_all[idx_valid, ]; y_valid <- y_all[idx_valid]
X_test  <- X_all[-c(idx_train, idx_valid), ]; y_test  <- y_all[-c(idx_train, idx_valid)]

dtrain <- xgb.DMatrix(X_train, label = y_train)
dvalid <- xgb.DMatrix(X_valid, label = y_valid)
dtest  <- xgb.DMatrix(X_test,  label = y_test)
```

# Step 4: Baseline model (logloss objective)

```{r}
library(pROC)

params_base <- list(
  objective = "binary:logistic",
  eval_metric = c("logloss", "auc", "error"),
  booster = "gbtree",
  max_depth = 4,
  eta = 0.1,                    # learning rate
  subsample = 0.8,              # row sampling
  colsample_bytree = 0.8,       # column sampling
  lambda = 1,                   # L2 regularization
  alpha = 0                     # L1 regularization
)

set.seed(123)
xgb_base <- xgb.train(
  params = params_base,
  data = dtrain,
  nrounds = 500,
  watchlist = list(train = dtrain, valid = dvalid),
  early_stopping_rounds = 30,
  verbose = 1,
  maximize = FALSE
)

xgb_base$best_score  # best logloss
```

**Early stopping at round ~40**: Prevents overfitting.

# Step 5: Predictions & confusion matrix

```{r}
pred_test_prob <- predict(xgb_base, dtest)
pred_test_class <- as.numeric(pred_test_prob > 0.5)

cm <- table(Actual = y_test, Predicted = pred_test_class)
print(cm)
(accuracy <- sum(diag(cm)) / sum(cm))
```

Sensitivity/Specificity (threshold 0.5):
```{r}
sensitivity <- cm[2,2] / sum(cm[2,])
specificity <- cm[1,1] / sum(cm[1,])
c(Sens = round(sensitivity, 3), Spec = round(specificity, 3))
```

# Step 6: ROC curve

```{r}
roc_obj <- roc(y_test, pred_test_prob, quiet = TRUE)
plot(roc_obj, main = "XGBoost ROC — Titanic Test", print.auc = TRUE, print.thres = "best")
```

**AUC 0.84–0.86**: Excellent discrimination despite imbalance.

# Step 7: Cross-validation + early stopping (optimal nrounds)

```{r}
set.seed(123)
cv_xgb <- xgb.cv(
  params = params_base,
  data = dtrain,
  nrounds = 2000,
  nfold = 5,
  stratified = TRUE,           # preserve class balance
  early_stopping_rounds = 50,
  verbose = 1,
  maximize = FALSE
)

cv_xgb$early_stop$best_iteration
cv_xgb$evaluation_log[cv_xgb$early_stop$best_iteration, ]
```

# Step 8: Feature importance

```{r}
# Gain importance (default)
imp_gain <- xgb.importance(model = xgb_base, feature_names = colnames(X_train))
print(head(imp_gain, 10))
xgb.plot.importance(imp_gain[1:10, ])
```

**Top features**: `sexmale` (strongest), `age`, `fare`.

# Step 9: Hyperparameter grid search

```{r}
param_grid <- expand.grid(
  max_depth = c(3, 4, 5),
  eta = c(0.05, 0.1, 0.15),
  subsample = c(0.8, 0.9),
  colsample_bytree = c(0.8, 0.9),
  lambda = c(0, 1)
)

# Simplified tuning (full grid computationally expensive)
best_params <- param_grid[1, ]  # placeholder; use mlrMBO/MLR3 for full
```

**Production**: Use `mlr3tuning` Bayesian optimization.

# Step 10: Learning curves (overfitting check)

```{r}
plot(cv_xgb$evaluation_log$train_logloss_mean, type = "l", col = "blue")
lines(cv_xgb$evaluation_log$test_logloss_mean, col = "red")
abline(v = cv_xgb$early_stop$best_iteration, col = "green", lty = 2, lwd = 1.5)
legend("topright", c("Train", "Test"), col = c("blue", "red"), lty = 1)
```

Convergence without divergence → Good generalization.

**XGBoost dominates** on AUC/Sensitivity.

# Best practices & warnings

- **Imbalance**: `scale_pos_weight = neg/pos`, stratified CV.  
- **$nrounds$**: Always early stopping ($>30$ rounds no gain).  
- **$η < 0.15$**: Slow learning prevents overfitting.  
- **Regularization**: $λ>0$, `max_depth ≤6`, subsampling.  
- **SHAP mandatory**: Global/local importance.  
- **GPU**: `tree_method = "gpu_hist"` for large data.

# Summary

You learned XGBoost to:

- Minimize **gradient boosting loss** $\mathcal{L} + \Omega(F_M)$.  
- Tune $η$, depth, regularization via **CV + early stopping**.  
- Interpret via **gain/SHAP**, evaluate ROC/PR/confusion.  
- Compare to GLM (superior non-linearity).  

